---
title: "<span style='font-size: 30px;'>Tutorial: using scLASER to simulate longitudinal single-cell data with clinical outcome association (e.g., treatment response)</span>"
author: |
  [Lab Website - https://fanzhanglab.org](https://fanzhanglab.org)  
  [Lab GitHub - https://github.com/fanzhanglab](https://github.com/fanzhanglab)
date: "`r format(Sys.Date())`"
output:
  rmarkdown::html_document:
    code_folding: show
    theme:
      version: 5
      bootswatch: flatly
      primary: "#984EA3"
      base_font: {google: "Inter"}
      heading_font: {google: "Inter"}
      code_font: {google: "Fira Code"}
vignette: >
  %\VignetteIndexEntry{scLASER • Data Simulation }
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  warning  = FALSE,
  message  = FALSE,
  fig.width  = 7.2,
  fig.height = 4.8
)
```

```{r}
library(scLASER)
library(dplyr)
library(tidyr)
library(ggplot2)
```


### <span style="font-size:25px;">1. Simulate a scenario that captures the longitudinal evolution of interacting cell types across visits and their divergence between groups.</span>


Longitudinal changes in cell-type abundance are specified using **effect design matrices** that explicitly encode how each cell type evolves across visits and differs between groups.

In this framework:

- **Rows correspond to cell types** (e.g., A, B, C, …).
- **Columns correspond to visits** (V0, V1, V2, …).
- **Matrix entries represent proportional changes** relative to baseline.
  - Positive values indicate enrichment.
  - Negative values indicate depletion.
  - A value of `0` indicates no change.

Two matrices are provided:

- `effect_mat_progressor` — effects for individuals with `disease = 1`
- `effect_mat_control` — effects for individuals with `disease = 0`

---

### <span style="font-size:25px;">2. Simulate a two-timepoint case consisting of baseline (V0) and follow-up (V1).</span>

```{r}
cell_types  <- LETTERS[1:(7 + 3)]  # A-J
time_points <- 2                   # V0, V1

E_prog <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_ctrl <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_prog["A", ] <- c(0.00, +0.30)
E_prog["B", ] <- c(0.00, -0.30)
E_prog["I", ] <- c(0.00, +0.30)
E_prog["J", ] <- c(0.00, -0.30)

set.seed(2020)
d2 <- generate_dummy_data(
  n_cells = 150,
  sd_celltypes = 0.1,
  n_major_cell_types = 7,
  n_minor_cell_types = 3,
  relative_abundance = 0.4,
  n_major_interact_celltypes = 2,
  n_minor_interact_celltypes = 2,
  n_individuals = 30,
  n_batchs = 4,
  interaction_feature = "visit",
  time_points = 2,
  test_var = "disease",
  prop_disease = 0.5,
  seed = 2020,
  effect_mat_progressor = E_prog,
  effect_mat_control    = E_ctrl
)

obj2 <- scLASER(metadata = d2)
plot_celltype_proportions(
  obj2,
  disease_var = "disease",
  disease_labels = c("0" = "Non-responder", "1" = "Responder"),
  disease_colors = c("Non-responder" = "#4DAF4A", "Responder" = "#984EA3"), highlight_cell_types = NULL
)

E_prog

E_ctrl
```

---

### <span style="font-size:25px;"> 3. Simulate a three-timepoint case (V0, V1, V2)</span>


```{r}
time_points <- 3

E_prog <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_ctrl <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_prog["A", ] <- c(0.00, 0.30, 0.60)
E_prog["B", ] <- c(0.00, -0.30, -0.60)
E_prog["I", ] <- c(0.00, 0.30, 0.60)
E_prog["J", ] <- c(0.00, -0.30, -0.60)

d3 <- generate_dummy_data(
  n_cells = 150,
  sd_celltypes = 0.1,
  n_major_cell_types = 7,
  n_minor_cell_types = 3,
  relative_abundance = 0.4,
  n_major_interact_celltypes = 2,
  n_minor_interact_celltypes = 2,
  n_individuals = 30,
  n_batchs = 4,
  interaction_feature = "visit",
  time_points = time_points,
  test_var = "disease",
  prop_disease = 0.5,
  seed = 2020,
  effect_mat_progressor = E_prog,
  effect_mat_control    = E_ctrl
)

obj3 <- scLASER(metadata = d3)
plot_celltype_proportions(
  obj3,
  disease_var = "disease",
  disease_labels = c("0" = "Non-responder", "1" = "Responder"),
  disease_colors = c("Non-responder" = "#4DAF4A", "Responder" = "#984EA3"), highlight_cell_types = NULL
)
E_prog

E_ctrl
```

---

### <span style="font-size:25px;">4. Simulate a five-timepoint case </span>

```{r}
time_points <- 5
traj <- c(0.00, 0.30, 0.60, 0.20, 0.40)

E_prog <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_ctrl <- matrix(
  0, nrow = length(cell_types), ncol = time_points,
  dimnames = list(cell_types, paste0("V", 0:(time_points - 1)))
)

E_prog[c("A","I"), ] <- matrix(traj,  nrow = 2, ncol = time_points, byrow = TRUE)
E_prog[c("B","J"), ] <- matrix(-traj, nrow = 2, ncol = time_points, byrow = TRUE)

d5 <- generate_dummy_data(
  n_cells = 150,
  sd_celltypes = 0.1,
  n_major_cell_types = 7,
  n_minor_cell_types = 3,
  relative_abundance = 0.4,
  n_major_interact_celltypes = 2,
  n_minor_interact_celltypes = 2,
  n_individuals = 30,
  n_batchs = 4,
  interaction_feature = "visit",
  time_points = time_points,
  test_var = "disease",
  prop_disease = 0.5,
  seed = 2020,
  effect_mat_progressor = E_prog,
  effect_mat_control    = E_ctrl
)

obj5 <- scLASER(metadata = d5)

E_prog

E_ctrl

plot_celltype_proportions(
  obj5,
  disease_var = "disease",
  disease_labels = c("0" = "Non-responder", "1" = "Responder"),
  disease_colors = c("Non-responder" = "#4DAF4A", "Responder" = "#984EA3"), highlight_cell_types = NULL)
```

---

### <span style="font-size:25px;">5. (Optional) Generate pseudo-PCs</span>

```{r}
pcs_matrix <- generate_pseudo_pcs_time(
  d3,
  n_pcs = 20,
  cluster_pcs = 1:20,
  disease_pcs = 0,
  sex_pcs = 0,
  age_pcs = 0,
  bmi_pcs = 0,
  batch_pcs = 0,
  interaction_pcs = 0,
  visit_pcs = 0,
  subject_pcs = 0,
  cluster_ratio = 0.8,
  disease_ratio = 0,
  sex_ratio = 0,
  age_ratio = 0,
  bmi_ratio = 0,
  batch_ratio = 0,
  interaction_ratio = 0,
  visit_ratio = 0,
  subject_ratio = 0,
  scale_factor = 5,
  cluster_col = "cell_type",
  sex_col = "sex",
  age_col = "age",
  bmi_col = "bmi",
  batch_col = "batch",
  disease_col = "disease",
  visit_col = "visit",
  seed = 1234
)
dim(pcs_matrix)
```
